Cache Oblivious B-trees

### 2011-12-09 paper.tex

The goal of modifying front compression is to maintain some of the space benefits to compression while reducing the high decoding cost and keeping the size of D small. Bender et al.  provide a modification that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\ll{D}\gg)$ after compression. The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters, add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means add $k_{i}$ to the stream in its uncompressed form. By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?
%todo: add another paragraph or two about the second iteration
\subsection{Operators}
The operations supported are

\subsection{Complexity}

>>>>

The goal of modifying front compression is to maintain some of the space benefits to compression while reducing the high decoding cost and keeping the size of D small. Bender et al.  provide a modification for static COSB-trees that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression. The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters, add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means add $k_{i}$ to the stream in its uncompressed form. By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?

However, the above modification falls short for anticipating insertions and deletions. The previous modification relies on each key being decoded within $c||k^{*}||$ elements of the compressed array, but this is no longer guaranteed when new keys can be inserted to the left of $k$. To make the locality-preserving front compression technique suitable for dynamic COSB-trees, Bender et al. modify the algorithm to include an extra copied prefix for each key. Consider the $3c||k||$ characters that would be to the left of a new key$k^{*}$ after insertion. Let $k'$ be the key at the end of this range. The idea is to prevent the insertion of $k^{*}$ from affecting any non-copied keys just to the right of $k'$, since this could potentially increase their decode cost. Compute the length $l$ of the largest common prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's in this range. $l$ represents the minimum length for any lcp between $k^{*}$ and $k'$, since the keys are stored in lexicographic order. If there is a copied prefix of length at least $l$ or a copied key within this range, insert $k^{*}$ as usual. Inserting $k^{*}$ will have no negative effect in this respect, since the decode length is effectively reset by the copied key or prefix like before. However, if inserting $k^{*}$ causes the minimum length of the lcp's in this range to increase, we need to fix the compression. To do this, we update the lcp of the key directly to the right of $k^{*}$ and store an extra copy of this lcp with $k^{*}$ itself. In this way, we can charge the work of copying extra prefixes for a new key to the appropriate number of elements before it. This number is determined by the length of the extra copied prefix stored with the key.

\subsection{Operators}
%todo: figure out how to do range queries from this paper somehow...
The operations supported are dictated by the Dictionary Matching problem. These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{}, and \Range{}. The behavior of each operation for dynamic COSB-trees is described below. \Insertkonly{} and \Delete{} are also supported but not described here.

\Search{}: works as follows for a given key $k$. Compute the hash value for $k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using $\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$. Index into the hashdata layer and perform a sequential scan across the PMA structure from $\rho_{p}$  to $\rho_{s}$ to find $\rho$. On a match with value $\rho'$, index into the keydata layer and decode $\rho'$ to obtain the resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not found during the sequential scan return false.

\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan in the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$ or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys in the keydata layer.

\Range{}: The paper doesn't say, but presumably this is very similar to \Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the end of the range. Hash the range keys and identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the hashdata layer across from  $\rho_{b,p}$ to  $\rho_{e,s}$. Only start noting the beginning of range when the first key with hash value after $\rho_{b}$, and stop once we reach the first key with hash value after $\rho_{e}$. Decode and return the set of keys within the given range in the keydata layer.

\subsection{Complexity}
\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability (w.h.p.). Traversing the centroid tree requires  $O(1+\log_{B}N)$ memory transfers, since the tree is built using$O(N/logN)$ keys and is balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of $k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers. Obtaining $k$ from the keydata layer requires no more memory transfers than the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost identical in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively. These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p.. Similarly, range queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers, where in this case $k'$ is the other end of the range and $Q$ is the set of keys within the given range.

\Insertkonly{} and \Delete{} w.h.p. require $O(1+\log_{B}N+\log^{2}N||k||/B)$ memory transfers. As mentioned previously, due to rebuilds updates have amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers. The rest of the cost for insertions and deletions is due to having to find the correct key to delete or location for insert, which can be seen from the cost of \Search{}.
